---
title: "TBI RNA-seq — 01 Sample QC"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5,
  out.dir = "results/analysis/01_sample_qc/"
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(pheatmap)
  library(RColorBrewer)
  library(patchwork)
  library(here)
})
dir.create("results/analysis/01_sample_qc", recursive = TRUE, showWarnings = FALSE)
```

## Scientific questions

1. Which samples have adequate sequencing depth (>10 M mapped reads)?
2. Are alignment rates consistently high (>80%) across batches and treatment groups?
3. Are duplication rates elevated in any sample subset (library quality concern)?
4. Do any samples cluster as outliers in the raw-count space?
5. Is there a batch-related difference in technical quality metrics?

---

## Load data

```{r load}
counts <- readRDS("results/counts/count_matrix.rds")
meta   <- readRDS("results/counts/metadata.rds")

stopifnot(identical(colnames(counts), meta$sample_id))
cat(sprintf("Loaded %d genes × %d samples\n", nrow(counts), ncol(counts)))
```

---

## Sequencing depth

```{r depth}
meta$total_counts <- colSums(counts)
depth_cutoff <- 10e6

p_depth <- ggplot(meta, aes(x = reorder(sample_id, total_counts),
                             y = total_counts / 1e6,
                             fill = batch)) +
  geom_col() +
  geom_hline(yintercept = depth_cutoff / 1e6, linetype = "dashed", colour = "red") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  labs(x = NULL, y = "Total counts (millions)",
       title = "Sequencing depth per sample",
       caption = "Red dashed line = 10 M read threshold") +
  theme_bw(base_size = 9)

ggsave("results/analysis/01_sample_qc/depth_barplot.pdf", p_depth,
       width = 8, height = max(4, nrow(meta) * 0.12))
p_depth
```

```{r depth-flag}
low_depth <- meta %>% filter(total_counts < depth_cutoff)
if (nrow(low_depth) > 0) {
  cat("Samples below depth threshold:\n")
  print(low_depth[, c("sample_id", "batch", "treatment", "total_counts")])
} else {
  cat("All samples pass the depth threshold.\n")
}
```

---

## Genes detected

```{r detection}
meta$genes_detected <- colSums(counts > 0)

ggplot(meta, aes(x = treatment, y = genes_detected, fill = batch)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  facet_grid(sex ~ genotype) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(y = "Genes detected (count > 0)",
       title = "Gene detection rate by experimental group") +
  theme_bw()
```

---

## Read MultiQC alignment summary

> **Note:** This section reads the MultiQC alignment TSV exported from
> `results/multiqc/aligned/multiqc_data/multiqc_star.txt`. Run
> `snakemake results/multiqc/aligned/multiqc_report.html` first.

```{r multiqc, eval=file.exists("results/multiqc/aligned/multiqc_data/multiqc_star.txt")}
mqc <- fread("results/multiqc/aligned/multiqc_data/multiqc_star.txt")
# Standardise sample column name
if ("Sample" %in% names(mqc)) setnames(mqc, "Sample", "sample_id")

mqc <- mqc %>%
  mutate(sample_id = str_remove(sample_id, "_Log\\.final$")) %>%
  inner_join(meta[, c("sample_id", "batch", "treatment", "sex", "genotype")],
             by = "sample_id")

# Unique mapping rate
p_map <- ggplot(mqc, aes(x = treatment,
                          y = uniquely_mapped_percent,
                          fill = batch)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
  geom_hline(yintercept = 80, linetype = "dashed", colour = "red") +
  facet_wrap(~sex) +
  scale_fill_brewer(palette = "Set2") +
  labs(y = "Uniquely mapped reads (%)",
       title = "STAR unique alignment rate",
       caption = "Red dashed line = 80% threshold") +
  theme_bw()
print(p_map)

ggsave("results/analysis/01_sample_qc/alignment_rate.pdf", p_map,
       width = 8, height = 5)
```

---

## Outlier detection — hierarchical clustering of raw counts

```{r outlier-hclust}
# Use log2(CPM + 1) for distance-based clustering
cpm_mat <- log2(sweep(counts, 2, colSums(counts), "/") * 1e6 + 1)
sample_dist <- as.dist(1 - cor(cpm_mat, method = "spearman"))
hc <- hclust(sample_dist, method = "average")

# Colour bars
ann_col <- meta %>%
  column_to_rownames("sample_id") %>%
  select(batch, treatment, region, sex, genotype)

# Identify samples with correlation < 0.85 to all others (potential outliers)
cor_mat <- cor(cpm_mat, method = "spearman")
min_cor  <- apply(cor_mat, 1, function(x) min(x[x < 1]))
outliers <- names(min_cor[min_cor < 0.85])

if (length(outliers) > 0) {
  cat("Potential outlier samples (min Spearman r < 0.85):\n")
  print(outliers)
} else {
  cat("No outlier samples detected.\n")
}

pheatmap(cor_mat,
         annotation_col = ann_col,
         clustering_distance_rows = sample_dist,
         clustering_distance_cols = sample_dist,
         color = colorRampPalette(rev(brewer.pal(9, "RdYlBu")))(100),
         breaks = seq(0.7, 1, length.out = 101),
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "Sample-to-sample Spearman correlation",
         filename = "results/analysis/01_sample_qc/sample_correlation_heatmap.pdf",
         width = 10, height = 9)
```

---

## Summary and exclusion list

```{r summary}
qc_summary <- meta %>%
  mutate(
    pass_depth    = total_counts >= depth_cutoff,
    pass_outlier  = !sample_id %in% outliers
  ) %>%
  select(sample_id, batch, treatment, region, pmi, sex, genotype,
         total_counts, genes_detected, pass_depth, pass_outlier)

fail_samples <- qc_summary %>% filter(!pass_depth | !pass_outlier)

cat(sprintf("\n%d / %d samples flagged for review:\n",
            nrow(fail_samples), nrow(qc_summary)))
print(fail_samples)

write.csv(qc_summary,
          "results/analysis/01_sample_qc/qc_summary.csv",
          row.names = FALSE)

# Save list of samples to EXCLUDE (set manually after review)
# Adjust the filter below based on your assessment of flagged samples
exclude_ids <- fail_samples$sample_id
saveRDS(exclude_ids, "results/analysis/01_sample_qc/exclude_samples.rds")
cat("\nExclude list saved. Review and edit as needed before proceeding.\n")
```

---

## Session info
```{r session}
sessionInfo()
```
