---
title: "TBI RNA-seq — 02 PCA & Batch Correction"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 8, fig.height = 6
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(limma)
  library(variancePartition)
  library(ggrepel)
  library(patchwork)
  library(RColorBrewer)
  library(pheatmap)
})
dir.create("results/analysis/02_pca_batch", recursive = TRUE, showWarnings = FALSE)
set.seed(42)
```

## Scientific questions

1. What is the dominant source of transcriptome-wide variance — batch, treatment, sex, genotype, or brain region?
2. Does sequencing batch drive a major principal component, requiring correction in the DESeq2 model?
3. After VST normalisation, do biological groups (treatment, region) separate on PCA?
4. Is postmortem interval (PMI) confounded with any biological variable?
5. Do ipsilateral/contralateral pairs from the same animal cluster together, confirming paired structure?

---

## Load and filter data

```{r load}
counts     <- readRDS("results/counts/count_matrix.rds")
meta       <- readRDS("results/counts/metadata.rds")
exclude_ids <- readRDS("results/analysis/01_sample_qc/exclude_samples.rds")

if (length(exclude_ids) > 0) {
  cat("Excluding", length(exclude_ids), "QC-flagged samples\n")
  keep_idx <- !colnames(counts) %in% exclude_ids
  counts   <- counts[, keep_idx]
  meta     <- meta[keep_idx, ]
}

# Minimum-count filter: keep genes with ≥10 summed counts
keep_genes <- rowSums(counts) >= 10
counts     <- counts[keep_genes, ]
cat(sprintf("Retained %d genes × %d samples after filtering\n",
            nrow(counts), ncol(counts)))
```

---

## VST normalisation

```{r vst}
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ batch + treatment   # minimal design for normalisation
)
vst <- vst(dds, blind = TRUE)
vst_mat <- assay(vst)
```

---

## PCA — raw (batch not corrected)

```{r pca-raw}
pca_res <- prcomp(t(vst_mat), center = TRUE, scale. = FALSE)
pve     <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_df <- as.data.frame(pca_res$x[, 1:5]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta, by = "sample_id")

plot_pca <- function(df, x, y, colour_var, shape_var = NULL, title = NULL) {
  p <- ggplot(df, aes(.data[[x]], .data[[y]],
                      colour = .data[[colour_var]],
                      shape  = if (!is.null(shape_var)) .data[[shape_var]] else NULL)) +
    geom_point(size = 3, alpha = 0.8) +
    geom_text_repel(aes(label = sample_id), size = 2, max.overlaps = 15) +
    labs(x = sprintf("%s (%s%%)", x, pve[as.integer(sub("PC", "", x))]),
         y = sprintf("%s (%s%%)", y, pve[as.integer(sub("PC", "", y))]),
         title = title) +
    theme_bw()
  p
}

p1 <- plot_pca(pca_df, "PC1", "PC2", "batch",     title = "Batch")
p2 <- plot_pca(pca_df, "PC1", "PC2", "treatment", title = "Treatment")
p3 <- plot_pca(pca_df, "PC1", "PC2", "region",    title = "Brain region")
p4 <- plot_pca(pca_df, "PC1", "PC2", "sex",       title = "Sex")
p5 <- plot_pca(pca_df, "PC1", "PC2", "genotype",  title = "Genotype")
p6 <- plot_pca(pca_df, "PC1", "PC2", "pmi",       title = "PMI")

(p1 | p2) / (p3 | p4) / (p5 | p6)
ggsave("results/analysis/02_pca_batch/pca_raw_grid.pdf", width = 14, height = 18)
```

---

## Variance partitioning

```{r varpart}
# Quantify % variance explained by each experimental factor
form <- ~ (1 | batch) + (1 | treatment) + (1 | region) +
           (1 | sex)  + (1 | genotype)  + (1 | pmi)
vp <- fitExtractVarPartModel(vst_mat, form, meta)
plotVarPart(sortCols(vp),
            main = "Variance explained by each experimental factor") +
  theme_bw()
ggsave("results/analysis/02_pca_batch/variance_partition.pdf", width = 8, height = 5)
```

---

## Batch correction (limma::removeBatchEffect)

This corrected matrix is **for visualisation only**.
DESeq2 models (steps 03–04) include `batch` in the design formula instead.

```{r batch-correct}
vst_corrected <- removeBatchEffect(
  vst_mat,
  batch       = meta$batch,
  design      = model.matrix(~ treatment + region + sex + genotype + pmi, data = meta)
)

pca_corr <- prcomp(t(vst_corrected), center = TRUE, scale. = FALSE)
pve_corr <- round(100 * pca_corr$sdev^2 / sum(pca_corr$sdev^2), 1)

pca_corr_df <- as.data.frame(pca_corr$x[, 1:5]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta, by = "sample_id")

q1 <- plot_pca(pca_corr_df, "PC1", "PC2", "batch",     title = "Batch (corrected)")
q2 <- plot_pca(pca_corr_df, "PC1", "PC2", "treatment", title = "Treatment (corrected)")
q3 <- plot_pca(pca_corr_df, "PC1", "PC2", "region",    title = "Region (corrected)")
q4 <- plot_pca(pca_corr_df, "PC1", "PC2", "sex",       title = "Sex (corrected)")

(q1 | q2) / (q3 | q4)
ggsave("results/analysis/02_pca_batch/pca_corrected_grid.pdf", width = 12, height = 10)
```

---

## PC1–PC5 scree and metadata associations

```{r pc-assoc}
pcs <- pca_corr_df %>% select(PC1:PC5, treatment, region, pmi, sex, genotype, batch)

assoc_results <- lapply(c("treatment", "region", "pmi", "sex", "genotype", "batch"),
  function(var) {
    lapply(paste0("PC", 1:5), function(pc) {
      f <- aov(as.formula(paste(pc, "~", var)), data = pcs)
      data.frame(PC = pc, variable = var,
                 p_value = summary(f)[[1]][["Pr(>F)"]][1])
    }) %>% bind_rows()
  }) %>% bind_rows()

assoc_wide <- assoc_results %>%
  mutate(neg_log10_p = -log10(p_value)) %>%
  pivot_wider(id_cols = variable, names_from = PC, values_from = neg_log10_p)

pheatmap(
  as.matrix(assoc_wide[, -1]),
  labels_row  = assoc_wide$variable,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "#d73027"))(50),
  main = "PC–metadata association (-log10 ANOVA p)",
  filename = "results/analysis/02_pca_batch/pc_metadata_assoc.pdf",
  width = 6, height = 4
)
```

---

## Save corrected matrix for downstream use

```{r save}
saveRDS(vst_mat,       "results/counts/vst_matrix.rds")
saveRDS(vst_corrected, "results/counts/vst_corrected_matrix.rds")
saveRDS(meta,          "results/counts/metadata_filtered.rds")
cat("VST and batch-corrected matrices saved.\n")
```

---

## Session info
```{r session}
sessionInfo()
```
