---
title: "TBI RNA-seq — 03 DESeq2 Main Contrasts"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(apeglm)
  library(EnhancedVolcano)
  library(pheatmap)
  library(RColorBrewer)
  library(patchwork)
  library(ggrepel)
  library(ComplexHeatmap)
})
dir.create("results/analysis/03_deseq2_main", recursive = TRUE, showWarnings = FALSE)
set.seed(42)

FDR <- 0.05
LFC <- 1.0   # |log2FoldChange| threshold for labelling
```

## Scientific questions

| Contrast | Scientific question |
|---|---|
| TBI vs SHAM | What transcriptional programme is activated by traumatic brain injury? |
| TBI.EGT vs TBI | Does EGT drug treatment suppress injury-induced gene expression? |
| TBI.EGT vs SHAM | Does EGT restore the transcriptome toward the uninjured baseline? |
| Ips vs Con | How does the injured hemisphere differ transcriptionally from the contralateral side? |
| PMI 24h vs 8h | How does time of sacrifice affect gene expression? (biological signal vs degradation artefact) |
| Male vs Female | Are there sex-biased transcriptional responses to TBI? |
| KO vs WT | How does the KO genotype modify gene expression under these conditions? |

---

## Load data

```{r load}
counts <- readRDS("results/counts/count_matrix.rds")
meta   <- readRDS("results/counts/metadata_filtered.rds")  # QC-filtered metadata
counts <- counts[, meta$sample_id]   # align columns

keep   <- rowSums(counts) >= 10
counts <- counts[keep, ]
cat(sprintf("%d genes × %d samples\n", nrow(counts), ncol(counts)))
```

---

## Helper functions

```{r helpers}
# Run DESeq2 for a given design, return shrunken results for one contrast
run_deseq2 <- function(counts, meta, design_formula, contrast_arg,
                        coef_name = NULL, result_name = NULL) {
  dds <- DESeqDataSetFromMatrix(
    countData = counts,
    colData   = meta,
    design    = design_formula
  )
  dds <- DESeq(dds, test = "Wald", fitType = "parametric",
               parallel = FALSE)

  if (!is.null(coef_name)) {
    # apeglm shrinkage (requires coef name)
    res <- lfcShrink(dds, coef = coef_name, type = "apeglm")
  } else {
    res <- results(dds, contrast = contrast_arg, alpha = FDR)
    res <- lfcShrink(dds, contrast = contrast_arg, type = "ashr")
  }
  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    arrange(padj, desc(abs(log2FoldChange)))
}

# Volcano plot wrapper
volcano_plot <- function(res_df, title, fdr = FDR, lfc = LFC) {
  EnhancedVolcano(res_df,
    lab        = res_df$gene_id,
    x          = "log2FoldChange",
    y          = "padj",
    pCutoff    = fdr,
    FCcutoff   = lfc,
    title      = title,
    subtitle   = sprintf("FDR < %s | |LFC| > %s", fdr, lfc),
    labSize    = 3,
    pointSize  = 1.5,
    legendPosition = "bottom")
}

# Summarise DEG counts
count_degs <- function(res_df, label, fdr = FDR, lfc = LFC) {
  sig <- res_df %>% filter(!is.na(padj) & padj < fdr & abs(log2FoldChange) >= lfc)
  data.frame(
    contrast = label,
    total    = nrow(sig),
    up       = sum(sig$log2FoldChange > 0),
    down     = sum(sig$log2FoldChange < 0)
  )
}

# Top-N heatmap of DEGs
deg_heatmap <- function(res_df, vst_mat, meta, top_n = 50,
                         fdr = FDR, lfc = LFC, title = "") {
  sig <- res_df %>%
    filter(!is.na(padj) & padj < fdr & abs(log2FoldChange) >= lfc) %>%
    slice_min(padj, n = top_n)
  if (nrow(sig) == 0) return(invisible(NULL))

  mat  <- vst_mat[sig$gene_id, , drop = FALSE]
  mat  <- t(scale(t(mat)))   # z-score per gene
  ann  <- meta %>%
    column_to_rownames("sample_id") %>%
    select(treatment, region, sex, genotype, batch)

  pheatmap(mat,
           annotation_col = ann,
           show_rownames  = nrow(mat) <= 60,
           show_colnames  = FALSE,
           color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
           breaks = seq(-3, 3, length.out = 101),
           main = title,
           fontsize_row = 7)
}

vst_mat <- readRDS("results/counts/vst_matrix.rds")
```

---

## Model 1 — Treatment effect (all samples)

**Design:** `~ batch + treatment`

```{r treatment-model}
dds_trt <- DESeqDataSetFromMatrix(counts, meta, ~ batch + treatment)
dds_trt <- DESeq(dds_trt)
resultsNames(dds_trt)
```

### TBI vs SHAM

```{r tbi-vs-sham}
res_tbi_sham <- lfcShrink(dds_trt,
  contrast = c("treatment", "TBI", "SHAM"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_tbi_sham, "results/analysis/03_deseq2_main/TBI_vs_SHAM.csv",
          row.names = FALSE)

volcano_plot(res_tbi_sham, "TBI vs SHAM")
ggsave("results/analysis/03_deseq2_main/volcano_TBI_vs_SHAM.pdf", width = 8, height = 7)
deg_heatmap(res_tbi_sham, vst_mat, meta, title = "Top DEGs: TBI vs SHAM")
```

### EGT vs TBI (drug effect)

```{r egt-vs-tbi}
res_egt_tbi <- lfcShrink(dds_trt,
  contrast = c("treatment", "TBI.EGT", "TBI"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_egt_tbi, "results/analysis/03_deseq2_main/EGT_vs_TBI.csv",
          row.names = FALSE)

volcano_plot(res_egt_tbi, "TBI.EGT vs TBI (drug effect)")
ggsave("results/analysis/03_deseq2_main/volcano_EGT_vs_TBI.pdf", width = 8, height = 7)
deg_heatmap(res_egt_tbi, vst_mat, meta, title = "Top DEGs: EGT vs TBI")
```

### EGT vs SHAM (injury recovery)

```{r egt-vs-sham}
res_egt_sham <- lfcShrink(dds_trt,
  contrast = c("treatment", "TBI.EGT", "SHAM"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_egt_sham, "results/analysis/03_deseq2_main/EGT_vs_SHAM.csv",
          row.names = FALSE)

volcano_plot(res_egt_sham, "TBI.EGT vs SHAM (residual injury)")
ggsave("results/analysis/03_deseq2_main/volcano_EGT_vs_SHAM.pdf", width = 8, height = 7)
```

---

## Overlap: genes rescued by EGT

```{r rescued}
sig_tbi_sham <- res_tbi_sham %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
sig_egt_tbi  <- res_egt_tbi  %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
sig_egt_sham <- res_egt_sham %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)

# Rescued = significant in TBI vs SHAM AND reversed in EGT vs TBI
rescued_up <- intersect(
  sig_tbi_sham$gene_id[sig_tbi_sham$log2FoldChange > 0],
  sig_egt_tbi$gene_id [sig_egt_tbi$log2FoldChange  < 0]
)
rescued_dn <- intersect(
  sig_tbi_sham$gene_id[sig_tbi_sham$log2FoldChange < 0],
  sig_egt_tbi$gene_id [sig_egt_tbi$log2FoldChange  > 0]
)

cat(sprintf("EGT-rescued genes (TBI-upregulated, reversed by EGT): %d\n", length(rescued_up)))
cat(sprintf("EGT-rescued genes (TBI-downregulated, restored by EGT): %d\n", length(rescued_dn)))

rescued_df <- data.frame(
  gene_id  = c(rescued_up, rescued_dn),
  direction = c(rep("rescued_up", length(rescued_up)),
                rep("rescued_down", length(rescued_dn)))
)
write.csv(rescued_df, "results/analysis/03_deseq2_main/EGT_rescued_genes.csv",
          row.names = FALSE)
```

---

## Model 2 — Brain region effect

**Design:** `~ batch + region` (pooling all treatments for overall regional effect)

```{r region-model}
dds_reg <- DESeqDataSetFromMatrix(counts, meta, ~ batch + region)
dds_reg <- DESeq(dds_reg)

res_ips_con <- lfcShrink(dds_reg,
  contrast = c("region", "Ips", "Con"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_ips_con, "results/analysis/03_deseq2_main/Ips_vs_Con.csv",
          row.names = FALSE)

volcano_plot(res_ips_con, "Ipsilateral vs Contralateral (all treatments)")
ggsave("results/analysis/03_deseq2_main/volcano_Ips_vs_Con.pdf", width = 8, height = 7)
```

---

## Model 3 — Postmortem interval

**Design:** `~ batch + pmi`
**Question:** Does 24h sacrifice vs 8h introduce transcript-stability artefacts?

```{r pmi-model}
dds_pmi <- DESeqDataSetFromMatrix(counts, meta, ~ batch + pmi)
dds_pmi <- DESeq(dds_pmi)

res_pmi <- lfcShrink(dds_pmi,
  contrast = c("pmi", "24h", "8h"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_pmi, "results/analysis/03_deseq2_main/PMI_24h_vs_8h.csv",
          row.names = FALSE)

volcano_plot(res_pmi, "PMI 24h vs 8h")
ggsave("results/analysis/03_deseq2_main/volcano_PMI.pdf", width = 8, height = 7)
```

---

## Model 4 — Sex

**Design:** `~ batch + sex`

```{r sex-model}
dds_sex <- DESeqDataSetFromMatrix(counts, meta, ~ batch + sex)
dds_sex <- DESeq(dds_sex)

res_sex <- lfcShrink(dds_sex,
  contrast = c("sex", "Male", "Female"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_sex, "results/analysis/03_deseq2_main/Male_vs_Female.csv",
          row.names = FALSE)

volcano_plot(res_sex, "Male vs Female (all conditions)")
ggsave("results/analysis/03_deseq2_main/volcano_sex.pdf", width = 8, height = 7)
```

---

## Model 5 — Genotype

**Design:** `~ batch + genotype`

```{r geno-model}
dds_geno <- DESeqDataSetFromMatrix(counts, meta, ~ batch + genotype)
dds_geno <- DESeq(dds_geno)

res_geno <- lfcShrink(dds_geno,
  contrast = c("genotype", "KO", "WT"), type = "ashr") %>%
  as.data.frame() %>% rownames_to_column("gene_id") %>%
  arrange(padj, desc(abs(log2FoldChange)))

write.csv(res_geno, "results/analysis/03_deseq2_main/KO_vs_WT.csv",
          row.names = FALSE)

volcano_plot(res_geno, "KO vs WT (all conditions)")
ggsave("results/analysis/03_deseq2_main/volcano_genotype.pdf", width = 8, height = 7)
```

---

## DEG summary table

```{r deg-summary}
deg_summary <- bind_rows(
  count_degs(res_tbi_sham,  "TBI vs SHAM"),
  count_degs(res_egt_tbi,   "EGT vs TBI"),
  count_degs(res_egt_sham,  "EGT vs SHAM"),
  count_degs(res_ips_con,   "Ips vs Con"),
  count_degs(res_pmi,       "PMI 24h vs 8h"),
  count_degs(res_sex,       "Male vs Female"),
  count_degs(res_geno,      "KO vs WT")
)

knitr::kable(deg_summary, caption = sprintf("DEGs at FDR < %s, |LFC| ≥ %s", FDR, LFC))
write.csv(deg_summary, "results/analysis/03_deseq2_main/deg_summary.csv", row.names = FALSE)
```

---

## Session info
```{r session}
sessionInfo()
```
