---
title: "TBI RNA-seq — 04 DESeq2 Interaction Models"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(ggrepel)
  library(pheatmap)
  library(patchwork)
  library(RColorBrewer)
  library(EnhancedVolcano)
})
dir.create("results/analysis/04_deseq2_interactions", recursive = TRUE,
           showWarnings = FALSE)
set.seed(42)
FDR <- 0.05
LFC <- 1.0
```

## Scientific questions

| Model | Scientific question |
|---|---|
| Treatment × Sex | Do males and females mount different transcriptional responses to TBI and EGT? |
| Treatment × Genotype | Does the KO gene modify how the brain responds to injury or drug treatment? |
| Treatment × Region | Is the drug effect or injury response regionally asymmetric (Ips ≠ Con)? |
| Treatment × PMI | Does time of sacrifice confound the apparent treatment effect? |

---

## Load data

```{r load}
counts <- readRDS("results/counts/count_matrix.rds")
meta   <- readRDS("results/counts/metadata_filtered.rds")
counts <- counts[, meta$sample_id]
counts <- counts[rowSums(counts) >= 10, ]
vst_mat <- readRDS("results/counts/vst_matrix.rds")
cat(sprintf("%d genes × %d samples\n", nrow(counts), ncol(counts)))
```

---

## Helper: extract interaction terms

```{r helpers}
# Extract genes where the interaction term is significant
# i.e. the effect of treatment DIFFERS between levels of the moderator
extract_interaction <- function(dds, coef_pattern, result_name) {
  coefs <- resultsNames(dds)
  interaction_coefs <- grep(coef_pattern, coefs, value = TRUE)
  if (length(interaction_coefs) == 0) {
    warning("No coefficients matching: ", coef_pattern)
    return(NULL)
  }
  lapply(interaction_coefs, function(coef) {
    res <- lfcShrink(dds, coef = coef, type = "apeglm")
    as.data.frame(res) %>%
      rownames_to_column("gene_id") %>%
      mutate(coef = coef) %>%
      arrange(padj, desc(abs(log2FoldChange)))
  }) %>% bind_rows()
}
```

---

## Interaction 1: Treatment × Sex

**Design:** `~ batch + treatment + sex + treatment:sex`

Genes with significant `treatment:sex` interaction terms are those for which
the treatment effect differs between males and females.

```{r trt-sex}
dds_ts <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ batch + treatment + sex + treatment:sex
)
dds_ts$treatment <- relevel(dds_ts$treatment, ref = "SHAM")
dds_ts$sex       <- relevel(dds_ts$sex,       ref = "Female")
dds_ts <- DESeq(dds_ts)
resultsNames(dds_ts)

res_ts <- extract_interaction(dds_ts, "treatment.*\\.sex|sex.*\\.treatment",
                               "Treatment_x_Sex")
if (!is.null(res_ts)) {
  write.csv(res_ts,
            "results/analysis/04_deseq2_interactions/Treatment_x_Sex.csv",
            row.names = FALSE)

  # How many genes show sex-specific TBI response?
  sig_ts <- res_ts %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
  cat(sprintf("Genes with significant Treatment × Sex interaction: %d\n", nrow(sig_ts)))
}
```

### Sex-stratified volcano: TBI vs SHAM within each sex

```{r sex-stratified}
plot_sex_strat <- function(sex_level) {
  sub_counts <- counts[, meta$sex == sex_level]
  sub_meta   <- meta[meta$sex == sex_level, ]

  dds_s <- DESeqDataSetFromMatrix(sub_counts, sub_meta, ~ batch + treatment)
  dds_s <- DESeq(dds_s)

  res <- lfcShrink(dds_s, contrast = c("treatment", "TBI", "SHAM"),
                   type = "ashr") %>%
    as.data.frame() %>% rownames_to_column("gene_id")
  write.csv(res,
    sprintf("results/analysis/04_deseq2_interactions/TBI_vs_SHAM_%s.csv", sex_level),
    row.names = FALSE)
  res
}

res_tbi_female <- plot_sex_strat("Female")
res_tbi_male   <- plot_sex_strat("Male")

# Scatter: male vs female LFC for TBI vs SHAM genes
lfc_compare <- inner_join(
  res_tbi_female %>% select(gene_id, lfc_female = log2FoldChange, padj_female = padj),
  res_tbi_male   %>% select(gene_id, lfc_male   = log2FoldChange, padj_male   = padj),
  by = "gene_id"
) %>%
  mutate(sex_specific = case_when(
    padj_female < FDR & padj_male >= FDR  ~ "Female-specific",
    padj_female >= FDR & padj_male < FDR  ~ "Male-specific",
    padj_female < FDR & padj_male < FDR   ~ "Shared",
    TRUE                                  ~ "NS"
  ))

ggplot(lfc_compare, aes(lfc_female, lfc_male, colour = sex_specific)) +
  geom_point(size = 1, alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
  scale_colour_manual(values = c("Female-specific" = "#e41a1c",
                                 "Male-specific"   = "#377eb8",
                                 "Shared"          = "#4daf4a",
                                 "NS"              = "grey80")) +
  labs(x = "log2FC (Female: TBI vs SHAM)",
       y = "log2FC (Male: TBI vs SHAM)",
       title = "Sex-stratified TBI response",
       colour = NULL) +
  theme_bw()
ggsave("results/analysis/04_deseq2_interactions/sex_LFC_scatter.pdf",
       width = 7, height = 6)
```

---

## Interaction 2: Treatment × Genotype

**Design:** `~ batch + treatment + genotype + treatment:genotype`

Tests whether the KO gene modifies the transcriptional response to TBI or EGT.

```{r trt-geno}
dds_tg <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ batch + treatment + genotype + treatment:genotype
)
dds_tg$treatment <- relevel(dds_tg$treatment, ref = "SHAM")
dds_tg$genotype  <- relevel(dds_tg$genotype,  ref = "WT")
dds_tg <- DESeq(dds_tg)
resultsNames(dds_tg)

res_tg <- extract_interaction(dds_tg, "treatment.*\\.genotype|genotype.*\\.treatment",
                               "Treatment_x_Genotype")
if (!is.null(res_tg)) {
  write.csv(res_tg,
            "results/analysis/04_deseq2_interactions/Treatment_x_Genotype.csv",
            row.names = FALSE)
  sig_tg <- res_tg %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
  cat(sprintf("Genes with significant Treatment × Genotype interaction: %d\n",
              nrow(sig_tg)))
}
```

### Genotype-stratified: TBI vs SHAM within WT and KO

```{r geno-stratified}
plot_geno_strat <- function(geno_level) {
  sub_counts <- counts[, meta$genotype == geno_level]
  sub_meta   <- meta[meta$genotype == geno_level, ]
  dds_g <- DESeqDataSetFromMatrix(sub_counts, sub_meta, ~ batch + treatment)
  dds_g <- DESeq(dds_g)
  res <- lfcShrink(dds_g, contrast = c("treatment", "TBI", "SHAM"),
                   type = "ashr") %>%
    as.data.frame() %>% rownames_to_column("gene_id")
  write.csv(res,
    sprintf("results/analysis/04_deseq2_interactions/TBI_vs_SHAM_%s.csv", geno_level),
    row.names = FALSE)
  res
}

res_tbi_wt <- plot_geno_strat("WT")
res_tbi_ko <- plot_geno_strat("KO")

# Genes that are specifically dysregulated in KO after TBI
ko_specific_up <- setdiff(
  res_tbi_ko$gene_id[!is.na(res_tbi_ko$padj) &
                     res_tbi_ko$padj < FDR &
                     res_tbi_ko$log2FoldChange > LFC],
  res_tbi_wt$gene_id[!is.na(res_tbi_wt$padj) &
                     res_tbi_wt$padj < FDR]
)
cat(sprintf("KO-specific TBI-upregulated genes: %d\n", length(ko_specific_up)))
```

---

## Interaction 3: Treatment × Region

**Design:** `~ batch + treatment + region + treatment:region`

Tests whether the ipsilateral vs contralateral difference is treatment-dependent.

```{r trt-region}
dds_tr <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ batch + treatment + region + treatment:region
)
dds_tr$treatment <- relevel(dds_tr$treatment, ref = "SHAM")
dds_tr$region    <- relevel(dds_tr$region,    ref = "Con")
dds_tr <- DESeq(dds_tr)
resultsNames(dds_tr)

res_tr <- extract_interaction(dds_tr, "treatment.*\\.region|region.*\\.treatment",
                               "Treatment_x_Region")
if (!is.null(res_tr)) {
  write.csv(res_tr,
            "results/analysis/04_deseq2_interactions/Treatment_x_Region.csv",
            row.names = FALSE)
  sig_tr <- res_tr %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
  cat(sprintf("Genes with significant Treatment × Region interaction: %d\n",
              nrow(sig_tr)))
}
```

---

## Interaction 4: Treatment × PMI

**Design:** `~ batch + treatment + pmi + treatment:pmi`

Tests whether the apparent treatment effect is confounded by time of sacrifice.

```{r trt-pmi}
dds_tp <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta,
  design    = ~ batch + treatment + pmi + treatment:pmi
)
dds_tp$treatment <- relevel(dds_tp$treatment, ref = "SHAM")
dds_tp$pmi       <- relevel(dds_tp$pmi,       ref = "8h")
dds_tp <- DESeq(dds_tp)
resultsNames(dds_tp)

res_tp <- extract_interaction(dds_tp, "treatment.*\\.pmi|pmi.*\\.treatment",
                               "Treatment_x_PMI")
if (!is.null(res_tp)) {
  write.csv(res_tp,
            "results/analysis/04_deseq2_interactions/Treatment_x_PMI.csv",
            row.names = FALSE)
  sig_tp <- res_tp %>% filter(!is.na(padj) & padj < FDR & abs(log2FoldChange) >= LFC)
  cat(sprintf("Genes with significant Treatment × PMI interaction: %d\n",
              nrow(sig_tp)))
}
```

---

## Session info
```{r session}
sessionInfo()
```
