---
title: "TBI RNA-seq — 05 Pathway Enrichment"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(clusterProfiler)
  library(enrichplot)
  library(ReactomePA)
  library(org.Mm.eg.db)
  library(fgsea)
  library(msigdbr)
  library(AnnotationDbi)
  library(patchwork)
  library(RColorBrewer)
})
dir.create("results/analysis/05_pathway", recursive = TRUE, showWarnings = FALSE)
set.seed(42)

FDR     <- 0.05
LFC     <- 1.0
P_PATH  <- 0.05   # pathway FDR threshold
```

## Scientific questions

1. What biological pathways (GO/KEGG/Reactome) are enriched in TBI vs SHAM DEGs?
2. Does EGT treatment target specific signalling pathways (neuroinflammation, apoptosis, synaptic plasticity)?
3. Are there sex-specific pathway responses to TBI?
4. Do KO-specific TBI DEGs implicate the known function of the KO gene?
5. What MSigDB Hallmarks are enriched across contrasts?

---

## Load DEG results

```{r load}
load_deg <- function(file) {
  read.csv(file) %>%
    filter(!is.na(padj) & !is.na(log2FoldChange))
}

deg_files <- list(
  TBI_vs_SHAM  = "results/analysis/03_deseq2_main/TBI_vs_SHAM.csv",
  EGT_vs_TBI   = "results/analysis/03_deseq2_main/EGT_vs_TBI.csv",
  EGT_vs_SHAM  = "results/analysis/03_deseq2_main/EGT_vs_SHAM.csv",
  Ips_vs_Con   = "results/analysis/03_deseq2_main/Ips_vs_Con.csv",
  PMI_24h_vs_8h = "results/analysis/03_deseq2_main/PMI_24h_vs_8h.csv",
  Sex_MvF      = "results/analysis/03_deseq2_main/Male_vs_Female.csv",
  Genotype_KOvWT = "results/analysis/03_deseq2_main/KO_vs_WT.csv"
)

degs <- lapply(deg_files[file.exists(unlist(deg_files))], load_deg)
cat("Loaded contrasts:", paste(names(degs), collapse = ", "), "\n")
```

---

## Gene ID conversion (Ensembl → Entrez)

```{r id-map}
# Build a mapping table once and reuse
all_genes <- unique(unlist(lapply(degs, `[[`, "gene_id")))
# Strip version suffix if present (ENSMUSG00000000001.8 → ENSMUSG00000000001)
all_genes_clean <- sub("\\.\\d+$", "", all_genes)

id_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys    = all_genes_clean,
  columns = c("ENTREZID", "SYMBOL", "GENENAME"),
  keytype = "ENSEMBL"
) %>%
  filter(!is.na(ENTREZID)) %>%
  distinct(ENSEMBL, .keep_all = TRUE)

add_entrez <- function(df) {
  df %>%
    mutate(ENSEMBL = sub("\\.\\d+$", "", gene_id)) %>%
    left_join(id_map, by = "ENSEMBL")
}
degs <- lapply(degs, add_entrez)
```

---

## Helper functions

```{r helpers}
get_sig_entrez <- function(df, fdr = FDR, lfc = LFC) {
  df %>%
    filter(padj < fdr & abs(log2FoldChange) >= lfc & !is.na(ENTREZID)) %>%
    pull(ENTREZID) %>% unique()
}

# Background = all tested genes with Entrez ID
get_universe <- function(df) {
  df %>% filter(!is.na(ENTREZID)) %>% pull(ENTREZID) %>% unique()
}

# Ranked list for GSEA (all genes, ranked by -log10(pvalue) * sign(LFC))
get_ranked <- function(df) {
  df %>%
    filter(!is.na(ENTREZID) & !is.na(pvalue)) %>%
    mutate(rank_stat = -log10(pvalue + 1e-300) * sign(log2FoldChange)) %>%
    arrange(desc(rank_stat)) %>%
    { setNames(.$rank_stat, .$ENTREZID) }
}

run_ora <- function(sig_entrez, universe, label, outdir) {
  if (length(sig_entrez) < 5) {
    message(label, ": too few significant genes for ORA")
    return(NULL)
  }
  # GO Biological Process
  ego <- enrichGO(gene         = sig_entrez,
                  universe     = universe,
                  OrgDb        = org.Mm.eg.db,
                  ont          = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff = P_PATH,
                  readable     = TRUE)
  # KEGG
  ekegg <- enrichKEGG(gene         = sig_entrez,
                      universe     = universe,
                      organism     = "mmu",
                      pAdjustMethod = "BH",
                      pvalueCutoff = P_PATH)
  # Reactome
  ereact <- enrichPathway(gene         = sig_entrez,
                           universe     = universe,
                           organism     = "mouse",
                           pAdjustMethod = "BH",
                           pvalueCutoff = P_PATH,
                           readable     = TRUE)

  list(GO = ego, KEGG = ekegg, Reactome = ereact)
}
```

---

## ORA for each main contrast

```{r ora-loop}
ora_results <- list()

for (contrast_name in names(degs)) {
  df     <- degs[[contrast_name]]
  sig_e  <- get_sig_entrez(df)
  univ_e <- get_universe(df)

  cat(sprintf("\n%s: %d significant genes\n", contrast_name, length(sig_e)))
  res <- run_ora(sig_e, univ_e, contrast_name,
                 outdir = "results/analysis/05_pathway")
  if (is.null(res)) next
  ora_results[[contrast_name]] <- res

  # Save tables
  for (db in names(res)) {
    if (!is.null(res[[db]]) && nrow(as.data.frame(res[[db]])) > 0) {
      write.csv(as.data.frame(res[[db]]),
        file.path("results/analysis/05_pathway",
                  sprintf("%s_%s_ORA.csv", contrast_name, db)),
        row.names = FALSE)
    }
  }
}
```

---

## Dot plots — TBI vs SHAM top pathways

```{r dotplot-tbi}
if (!is.null(ora_results[["TBI_vs_SHAM"]])) {
  p_go    <- dotplot(ora_results$TBI_vs_SHAM$GO,   showCategory = 20) +
               ggtitle("GO:BP — TBI vs SHAM")
  p_kegg  <- dotplot(ora_results$TBI_vs_SHAM$KEGG, showCategory = 20) +
               ggtitle("KEGG — TBI vs SHAM")

  p_go | p_kegg
  ggsave("results/analysis/05_pathway/dotplot_TBI_vs_SHAM.pdf",
         width = 16, height = 8)
}
```

---

## GSEA with MSigDB Hallmarks

```{r gsea-hallmarks}
hallmarks <- msigdbr(species = "Mus musculus", category = "H") %>%
  select(gs_name, entrez_gene) %>%
  distinct()
hallmarks_list <- split(hallmarks$entrez_gene, hallmarks$gs_name)

gsea_results <- list()

for (contrast_name in names(degs)) {
  ranked <- get_ranked(degs[[contrast_name]])
  if (length(ranked) < 100) next

  fgsea_res <- fgseaMultilevel(
    pathways   = hallmarks_list,
    stats      = ranked,
    minSize    = 15,
    maxSize    = 500,
    nPermSimple = 10000
  )
  gsea_results[[contrast_name]] <- fgsea_res %>%
    arrange(padj) %>%
    filter(padj < P_PATH)

  write.csv(gsea_results[[contrast_name]],
    file.path("results/analysis/05_pathway",
              sprintf("%s_Hallmarks_GSEA.csv", contrast_name)),
    row.names = FALSE)
}
```

### Hallmark GSEA heatmap across contrasts

```{r gsea-heatmap}
if (length(gsea_results) > 0) {
  all_sig_paths <- unique(unlist(lapply(gsea_results, function(r) r$pathway)))

  nes_mat <- sapply(names(gsea_results), function(cn) {
    r <- gsea_results[[cn]]
    sapply(all_sig_paths, function(p) {
      idx <- match(p, r$pathway)
      if (is.na(idx)) 0 else r$NES[idx]
    })
  })
  rownames(nes_mat) <- all_sig_paths

  # Keep only pathways significant in ≥1 contrast
  keep_rows <- rowSums(nes_mat != 0) >= 1
  nes_mat   <- nes_mat[keep_rows, , drop = FALSE]

  if (nrow(nes_mat) > 1) {
    pheatmap(nes_mat,
             color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
             breaks = seq(-3, 3, length.out = 101),
             cluster_cols = FALSE,
             fontsize_row = 8,
             main = "MSigDB Hallmarks NES across contrasts",
             filename = "results/analysis/05_pathway/hallmarks_NES_heatmap.pdf",
             width = 10, height = max(6, nrow(nes_mat) * 0.2))
  }
}
```

---

## MSigDB C2 (curated gene sets) GSEA — TBI vs SHAM

```{r gsea-c2}
c2_sets <- msigdbr(species = "Mus musculus", category = "C2") %>%
  select(gs_name, entrez_gene) %>% distinct()
c2_list <- split(c2_sets$entrez_gene, c2_sets$gs_name)

if ("TBI_vs_SHAM" %in% names(degs)) {
  ranked_tbi <- get_ranked(degs$TBI_vs_SHAM)
  fgsea_c2 <- fgseaMultilevel(
    pathways    = c2_list,
    stats       = ranked_tbi,
    minSize     = 15,
    maxSize     = 500,
    nPermSimple = 10000
  ) %>% arrange(padj)

  top_c2 <- fgsea_c2 %>% filter(padj < P_PATH) %>% slice_min(padj, n = 30)
  write.csv(fgsea_c2 %>% filter(padj < P_PATH),
            "results/analysis/05_pathway/TBI_vs_SHAM_C2_GSEA.csv",
            row.names = FALSE)

  # GSEA enrichment plot for top pathway
  if (nrow(top_c2) > 0) {
    plotEnrichment(c2_list[[top_c2$pathway[1]]], ranked_tbi) +
      ggtitle(paste("Top C2 pathway:", top_c2$pathway[1]))
    ggsave("results/analysis/05_pathway/GSEA_top_C2_TBI.pdf", width = 7, height = 4)
  }
}
```

---

## Enrichment map — TBI vs SHAM GO results

```{r emapplot}
if (!is.null(ora_results[["TBI_vs_SHAM"]]) &&
    !is.null(ora_results$TBI_vs_SHAM$GO) &&
    nrow(as.data.frame(ora_results$TBI_vs_SHAM$GO)) >= 5) {

  ego_sim <- pairwise_termsim(ora_results$TBI_vs_SHAM$GO)
  emapplot(ego_sim, showCategory = 30) +
    ggtitle("GO:BP enrichment map — TBI vs SHAM")
  ggsave("results/analysis/05_pathway/emapplot_TBI_vs_SHAM_GO.pdf",
         width = 12, height = 10)
}
```

---

## Session info
```{r session}
sessionInfo()
```
