---
title: "TBI RNA-seq — 06 Co-expression Network (WGCNA)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 10, fig.height = 7
)
suppressPackageStartupMessages({
  library(tidyverse)
  library(WGCNA)
  library(DESeq2)
  library(clusterProfiler)
  library(org.Mm.eg.db)
  library(AnnotationDbi)
  library(pheatmap)
  library(RColorBrewer)
  library(igraph)
})
options(stringsAsFactors = FALSE)
allowWGCNAThreads()
dir.create("results/analysis/06_network", recursive = TRUE, showWarnings = FALSE)
set.seed(42)
```

## Scientific questions

1. What co-expression modules are robustly associated with TBI severity or treatment?
2. Which hub genes are most central in TBI-associated modules, making them candidate targets or biomarkers?
3. Do co-expression modules reflect distinct cell types (neurons, astrocytes, microglia, endothelial)?
4. How does the KO genotype affect module structure?
5. Are PMI-associated modules distinct from injury-associated modules (separating biology from artefact)?

---

## Prepare input expression matrix

```{r load}
counts  <- readRDS("results/counts/count_matrix.rds")
meta    <- readRDS("results/counts/metadata_filtered.rds")
counts  <- counts[, meta$sample_id]

# Use variance-stabilised, batch-corrected matrix for WGCNA
vst_corr <- readRDS("results/counts/vst_corrected_matrix.rds")

# WGCNA requires samples as rows, genes as columns
expr_mat <- t(vst_corr)

cat(sprintf("Input: %d samples × %d genes\n", nrow(expr_mat), ncol(expr_mat)))
```

---

## Filter low-variance genes

```{r filter-var}
# Keep top 5000 most variable genes to reduce computation
gene_var <- apply(expr_mat, 2, var)
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:5000]
expr_mat  <- expr_mat[, top_genes]

cat(sprintf("After variance filtering: %d genes\n", ncol(expr_mat)))
```

---

## Sample network: detect outliers

```{r sample-outlier}
sft_tree <- hclust(dist(expr_mat), method = "average")
plot(sft_tree, labels = rownames(expr_mat),
     cex = 0.6, main = "Sample clustering (pre-WGCNA outlier detection)")
```

---

## Soft-thresholding power selection

```{r soft-threshold}
powers <- c(1:10, seq(12, 20, by = 2))
sft    <- pickSoftThreshold(expr_mat,
                             powerVector = powers,
                             networkType = "signed",
                             verbose     = 0)

# Plot scale-free topology fit
par(mfrow = c(1, 2))
plot(sft$fitIndices$Power,
     -sign(sft$fitIndices$slope) * sft$fitIndices$SFT.R.sq,
     xlab = "Soft threshold (power)", ylab = "Scale free topology R²",
     type = "b", main = "Scale-free topology fit")
abline(h = 0.85, col = "red", lty = 2)

plot(sft$fitIndices$Power, sft$fitIndices$mean.k.,
     xlab = "Soft threshold (power)", ylab = "Mean connectivity",
     type = "b", main = "Mean connectivity")

# Select power: smallest power with R² ≥ 0.85
soft_power <- sft$fitIndices$Power[
  which(-sign(sft$fitIndices$slope) * sft$fitIndices$SFT.R.sq >= 0.85)[1]
]
if (is.na(soft_power)) soft_power <- 12   # fallback
cat(sprintf("Selected soft-thresholding power: %d\n", soft_power))
```

---

## Network construction and module detection

```{r wgcna-build, cache=TRUE}
net <- blockwiseModules(
  expr_mat,
  power            = soft_power,
  networkType      = "signed",
  TOMType          = "signed",
  minModuleSize    = 30,
  mergeCutHeight   = 0.25,
  maxBlockSize     = 5000,
  numericLabels    = FALSE,
  saveTOMs         = FALSE,
  verbose          = 0
)

module_colours <- net$colors
cat(sprintf("Detected %d modules (excluding grey)\n",
            length(unique(module_colours[module_colours != "grey"]))))
table(module_colours) %>% sort(decreasing = TRUE)
```

---

## Module dendrogram

```{r module-dendro}
plotDendroAndColors(
  net$dendrograms[[1]],
  module_colours[net$blockGenes[[1]]],
  "Module colours",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05,
  main = "Gene dendrogram and module colours"
)
```

---

## Module eigengenes

```{r eigengenes}
MEs <- moduleEigengenes(expr_mat, module_colours)$eigengenes
MEs_ordered <- orderMEs(MEs)
```

---

## Module–trait correlation

```{r trait-corr}
# Build numeric trait matrix
trait_mat <- meta %>%
  select(treatment, region, sex, genotype, pmi, batch) %>%
  mutate(
    TBI        = as.integer(treatment == "TBI"),
    TBI_EGT    = as.integer(treatment == "TBI.EGT"),
    Ipsilateral = as.integer(region == "Ips"),
    Male       = as.integer(sex == "Male"),
    KO         = as.integer(genotype == "KO"),
    PMI_24h    = as.integer(pmi == "24h"),
    Batch2     = as.integer(batch == "B2")
  ) %>%
  select(TBI, TBI_EGT, Ipsilateral, Male, KO, PMI_24h, Batch2) %>%
  as.data.frame()
rownames(trait_mat) <- meta$sample_id

n_samples <- nrow(expr_mat)
module_trait_cor  <- cor(MEs_ordered, trait_mat, use = "pairwise.complete.obs")
module_trait_pval <- corPvalueFisher(module_trait_cor, n_samples)

# Heatmap
text_mat <- paste0(
  round(module_trait_cor, 2), "\n(",
  signif(module_trait_pval, 1), ")"
)
dim(text_mat) <- dim(module_trait_cor)

labeledHeatmap(
  Matrix     = module_trait_cor,
  xLabels    = colnames(trait_mat),
  yLabels    = names(MEs_ordered),
  ySymbols   = names(MEs_ordered),
  colorLabels = FALSE,
  colors     = blueWhiteRed(50),
  textMatrix = text_mat,
  setStdMargins = FALSE,
  cex.text   = 0.7,
  zlim       = c(-1, 1),
  main       = "Module–trait correlations (Pearson r, p-value)"
)

# Save
corr_df <- as.data.frame(module_trait_cor) %>%
  rownames_to_column("module")
write.csv(corr_df,
          "results/analysis/06_network/module_trait_correlations.csv",
          row.names = FALSE)
```

---

## Hub genes per module

```{r hub-genes}
# Module membership (kME): correlation of each gene with its module eigengene
kME <- cor(expr_mat, MEs_ordered, use = "pairwise.complete.obs")

# Top hub genes per module (highest |kME|)
hub_genes <- lapply(unique(module_colours[module_colours != "grey"]), function(mod) {
  mod_genes   <- names(module_colours)[module_colours == mod]
  me_col      <- paste0("ME", mod)
  if (!me_col %in% colnames(kME)) return(NULL)
  kme_vals    <- kME[mod_genes, me_col]
  top_hubs    <- names(sort(abs(kme_vals), decreasing = TRUE))[1:min(20, length(kme_vals))]
  data.frame(module = mod, gene_id = top_hubs,
             kME = kme_vals[top_hubs])
}) %>%
  bind_rows() %>%
  arrange(module, desc(abs(kME)))

# Map to gene symbols
hub_genes <- hub_genes %>%
  mutate(ENSEMBL = sub("\\.\\d+$", "", gene_id)) %>%
  left_join(
    AnnotationDbi::select(org.Mm.eg.db,
      keys    = unique(sub("\\.\\d+$", "", hub_genes$gene_id)),
      columns = c("SYMBOL", "GENENAME"),
      keytype = "ENSEMBL"),
    by = "ENSEMBL"
  )

write.csv(hub_genes, "results/analysis/06_network/hub_genes_per_module.csv",
          row.names = FALSE)
cat("Top 5 hub genes per module:\n")
hub_genes %>% group_by(module) %>% slice_head(n = 5) %>%
  select(module, SYMBOL, kME) %>% print(n = Inf)
```

---

## GO enrichment of TBI-associated modules

```{r module-go}
# Identify modules strongly correlated with TBI trait (|r| > 0.3, p < 0.05)
tbi_modules <- rownames(module_trait_cor)[
  abs(module_trait_cor[, "TBI"]) > 0.3 &
  module_trait_pval[, "TBI"] < 0.05
]
cat("TBI-associated modules:", paste(tbi_modules, collapse = ", "), "\n")

for (mod in gsub("ME", "", tbi_modules)) {
  mod_genes   <- sub("\\.\\d+$", "", names(module_colours)[module_colours == mod])
  universe    <- sub("\\.\\d+$", "", colnames(expr_mat))

  entrez_mod  <- AnnotationDbi::select(org.Mm.eg.db,
    keys = mod_genes, columns = "ENTREZID", keytype = "ENSEMBL") %>%
    filter(!is.na(ENTREZID)) %>% pull(ENTREZID)
  entrez_univ <- AnnotationDbi::select(org.Mm.eg.db,
    keys = universe, columns = "ENTREZID", keytype = "ENSEMBL") %>%
    filter(!is.na(ENTREZID)) %>% pull(ENTREZID)

  if (length(entrez_mod) < 10) next

  ego <- enrichGO(gene         = entrez_mod,
                  universe     = entrez_univ,
                  OrgDb        = org.Mm.eg.db,
                  ont          = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  readable     = TRUE)
  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
    p <- dotplot(ego, showCategory = 15) +
      ggtitle(sprintf("Module %s — GO:BP enrichment", mod))
    ggsave(sprintf("results/analysis/06_network/GO_module_%s.pdf", mod),
           p, width = 8, height = 6)
    write.csv(as.data.frame(ego),
      sprintf("results/analysis/06_network/GO_module_%s.csv", mod),
      row.names = FALSE)
  }
}
```

---

## Export for Cytoscape

```{r cytoscape-export}
# Export the top hub-gene network for the strongest TBI module
if (length(tbi_modules) > 0) {
  top_mod    <- gsub("ME", "", tbi_modules[1])
  mod_genes  <- names(module_colours)[module_colours == top_mod]

  # Keep top 50 hubs by kME
  top50 <- hub_genes %>%
    filter(module == top_mod) %>%
    slice_head(n = 50) %>%
    pull(gene_id)

  # Sub-TOM for these genes
  sub_expr <- expr_mat[, intersect(top50, colnames(expr_mat))]
  sub_tom  <- TOMsimilarity(adjacency(sub_expr, power = soft_power,
                                      type = "signed"),
                             TOMType = "signed")
  rownames(sub_tom) <- colnames(sub_tom) <- colnames(sub_expr)

  # Build edge list (weight > 0.05)
  edge_df <- as.data.frame(as.table(sub_tom)) %>%
    setNames(c("from", "to", "weight")) %>%
    filter(from < to & weight > 0.05) %>%
    arrange(desc(weight))

  write.csv(edge_df,
    sprintf("results/analysis/06_network/cytoscape_edges_module_%s.csv", top_mod),
    row.names = FALSE)

  node_df <- hub_genes %>%
    filter(module == top_mod, gene_id %in% top50) %>%
    select(gene_id, SYMBOL, kME)
  write.csv(node_df,
    sprintf("results/analysis/06_network/cytoscape_nodes_module_%s.csv", top_mod),
    row.names = FALSE)

  cat(sprintf("Exported %d edges and %d nodes for module %s → Cytoscape\n",
              nrow(edge_df), nrow(node_df), top_mod))
}
```

---

## Session info
```{r session}
sessionInfo()
```
